---
title: "Cheatsheet elements"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(vtree)
```

# Pruning revised

```{r}
(r1 <- vtree(FakeData,"Severity Sex",sameline=TRUE))
```

```{r}
grVizToPNG(r1,width=3000,folder="png")
```

```{r}
(r2 <- vtree(FakeData,"Severity Sex",sameline=TRUE,
  prune=list(Severity=c("Mild","Moderate"))))
```

```{r}
grVizToPNG(r2,width=3000,folder="png")
```

vtree(FakeData,"Group Category",sameline=TRUE)

```{r}
(r3 <- vtree(FakeData,"Category Group",sameline=TRUE))
```

```{r}
grVizToPNG(r3,width=3000,folder="png")
```

```{r}
(r4 <- vtree(FakeData,"Category Group",sameline=TRUE,
  prunebelow=list(Category=c("single","double"))))
```

```{r}
grVizToPNG(r4,width=3000,folder="png")
```

```{r}
(r5 <- vtree(FakeData,"Category Group",sameline=TRUE,
  keep=list(Category=c("single","double"))))
```

```{r}
grVizToPNG(r5,width=3000,folder="png")
```

```{r}
(r6 <- vtree(FakeData,"Category Group",sameline=TRUE,
  follow=list(Category=c("single","double"))))
```

```{r}
grVizToPNG(r6,width=3000,folder="png")
```

```{r}
(r7 <- vtree(FakeData,"Category Group",sameline=TRUE,showlegend=TRUE))
```

```{r}
grVizToPNG(r7,width=3000,folder="png")
```


# Basics

```{r}
(v1 <- vtree(FakeData,"Severity Sex",NAfillcolor="#FFFDFD"))
```

```{r}
grVizToPNG(v1,width=3000,folder="png")
```


```{r}
(v1a <- vtree(FakeData,"Severity Sex",squeeze=0.8))
```

```{r}
grVizToPNG(v1a,width=3000,folder="png")
```


```{r}
(v2 <- vtree(FakeData,"Group",horiz=FALSE))
```

```{r}
grVizToPNG(v2,width=3000,folder="png")
```

```{r}
(v3 <- vtree(FakeData,"Category Group",sameline=TRUE,
  showlegend=TRUE))
```

```{r}
grVizToPNG(v3,width=3000,folder="png")
```

## Variable specification modifiers

Modifier         | Effect
-----------------|-------------------------------------------------
prefix `is.na:`  | replace variable with missing value indicator
prefix `stem:`   | include all variables from REDCap with names based on stem
prefix `tri:`    | within each node, replace variable with trichotomized variable 
`variable=value` | dichotomize variable at value
`variable<value` | dichotomize variable below value
`variable>value` | dichotomize variable above value


# Pruning

Parameter     | Effect
--------------|---------------------------------------------------
`prune`       | remove all specified nodes
`prunebelow`  | remove all descendants of the specified nodes
`keep`        | retain only the specified nodes
`follow`      | retain descendants of only the specified nodes

```{r}
(v4 <- vtree(FakeData,"Severity Sex",
  prune=list(Severity=c("Mild","Moderate"))))
```

```{r}
grVizToPNG(v4,width=3000,folder="png")
```

```{r}
(v5 <- vtree(FakeData,"Severity Sex",
  prunebelow=list(Severity=c("Mild","Moderate"))))
```

```{r}
grVizToPNG(v5,width=3000,folder="png")
```

```{r}
(v6 <- vtree(FakeData,"Severity Sex",
  keep=list(Severity=c("Mild","NA"))))
```

```{r}
grVizToPNG(v6,width=3000,folder="png")
```

```{r}
(v7 <- vtree(FakeData,"Severity Sex",
  follow=list(Severity=c("Mild","NA"))))
```

```{r}
grVizToPNG(v7,width=3000,folder="png")
```

# Labeling

Parameter     | Example
--------------|----------------------------------------------
`labelvar`    | `labelvar=c(Severity="New label for Severity")`
`labelnode`   | `labelnode=list(Sex=c(Male="M",Female="F"))`
`tlabelnode`  | `tlabelnode=list(c(Group="A",Sex="F",label="girl"))`


# Text

Code                  | Effect
----------------------|--------------------------
`\n`                  | Line break
`*...*`               | Italics
`**...**`             | Bold
`^...^`               | Superscript
`~...~`               | Subscript
`%%red` ...`%%`       | Make text red (or whatever color)

```{r}
(v8 <- vtree(FakeData,"Group Category",sameline=TRUE,
  text=list(Category=c(triple="\n*not verified*"))))
```

```{r}
grVizToPNG(v8,width=3000,folder="png")
```

```{r}
(v9 <- vtree(FakeData,"Severity Age Pre Post",check.is.na=TRUE))
```

```{r}
grVizToPNG(v9,width=3000,folder="png")
```

# Summary

## Statistic codes

Code            | Result
:---------------|:-------------------------------------------------------------------
`%mean%`        |  mean
`%SD%`          |  standard deviation
`%min%`         |  minimum
`%max%`         |  maximum
`%pX%`          |  Xth percentile (e.g. `p50` means the 50th percentile)
`%median%`      |  median, i.e. p50
`%IQR%`         |  IQR, i.e. p25, p75
`%npct%`        |  frequency and percentage
`%pct%`         |  just percentage
`%list%`        |  list of individual values, separated by commas
`%listlines%`   |  list of individual values, each on a separate line
`%mv%`          |  the number of missing values
`%trunc=`*n*`%` |  flag: Truncate the summary to the first *n* characters. 

## Restriction codes

Code            | Summary information restricted to:
:---------------|:---------------------------------------
`%noroot%`      | all nodes *except* the root
`%leafonly%`    | leaf nodes
`%var=`*v*`%`   | nodes of variable *v*
`%node=`*n*`%`  | nodes named *n*


## Variable specification modifiers

Modifier         | Effect
-----------------|-------------------------------------------------
`variable=value` | dichotomize variable at value
`variable<value` | dichotomize variable below value
`variable>value` | dichotomize variable above value


```{r}
(v10 <- vtree(FakeData,"Severity",sameline=TRUE,
  summary="Score , mean=%mean%"))
```

```{r}
grVizToPNG(v10,width=3000,folder="png")
```
